Dissertation

### 2024-04-15 Manuscript/manuscript.qmd
few tweaks

A distinct sub-literature within the category learning domain has examined how the variability or dispersion of the categories themselves influences generalization to ambiguous regions of the category space (e.g. the region between the two categories). The general approach is to train participants with examples from a high variability category and a low variability category. Participants are then tested with novel items located within ambiguous regions of the category space which allow the experimenters to assess whether the difference in category variability influenced how far participants generalize the category boundaries. @cohenCategoryVariabilityExemplar2001 trained subjects on two categories, one with much more variability than the other. In experiment 1, a low variability category composed of 1 instance was compared against a high-variability category of 2 instances in one condition, and 7 instances in another. In experiment 2 both categories were composed of 3 instances, but for the low-variability group the instances were clustered close to each other, whereas the high-variability groups instances were spread much further apart. Participants were tested on an ambiguous novel instance that was located in between the two trained categories. Both experiments provided evidence that participants were much more likely to categorize the novel middle stimulus into a category with greater variation.

@@

Early models of learning often assumed that discrete experiences with some task or category were not stored individually in memory, but instead promoted the formation of a summary representation, often referred to as a prototype or schema, and that exposure to novel examples would then prompt the retrieval of whichever preexisting prototype was most similar. In addition to being a landmark study on the influence of training variability, @posnerGenesisAbstractIdeas1968 (described above) also put forward an influential argument concerning the nature of the mental representations acquired during learning - namely that learners tend to abstract a prototype, or aggregate representation of the dot pattern categories, rather than encoding each individual stimuli. Recall that participants are trained on only on distortions of the category prototypes (e.g.  low, medium or high distortions), never encountering the exact prototypes during the training stage. Then, in the testing phase, participants are tested with the prototype patterns, their old training items, and novel low, medium and high distortions. The authors found that participants had the highest testing accuracy for the previously unseen prototype patterns, followed by the old training items, and then the novel low, medium and high distortions. The authors interpreted this pattern as evidence that participants had acquired prototype representation of the category, as opposed to storing each individual training instance, and that generalization was based on the similarity of the testing items to the learned prototype representations. @posnerGenesisAbstractIdeas1968 has been extremely influential,  and continues to be cited in contemporary research as clear evidence that prototype abstraction underlies the benefits of varied training. It's also referenced as a key influence in the development of "Schema Theory of Motor Learning" @schmidtSchemaTheoryDiscrete1975, which in turn influenced decades of research on the potential benefits of varied training in motor skill learning. However a number of the core assumptions utilized by @posnerGenesisAbstractIdeas1968 were later called into question both empirically and with competing theoretical accounts [@hintzmanMINERVASimulationModel1984; @hintzmanSchemaAbstractionMultipletrace1986; @palmeriCentralTendenciesExtreme2001; @zakiHighdistortionEnhancementEffect2007; @mcclellandDistributedMemoryRepresentation1985; @nosofskyInvestigationsExemplarBasedConnectionist1992; @knappTheoryCategorizationBased1984]. @palmeriCentralTendenciesExtreme2001 demonstrated both the dangers of assuming that psychological representations mimic the metric stimulus space, as well the viability of models with simpler representational assumptions. These authors conducted a near replication of the @posnerGenesisAbstractIdeas1968 study, but also had participants provide similarity judgements of the dot pattern stimuli after completing the training phase. A multidimensional scaling analysis of the similarity judgements revelead that the psychological representations of the prototype stimuli were not located in the middle of the training stimuli, but were instead extreme points in the psychological space. The authors also demonstrated the generalization patterns of @posnerGenesisAbstractIdeas1968 could be accounted for by an exemplar-based model, without any need to assume the abstraction of a prototype.

>>>>

A distinct sub-literature within the category learning domain has examined how the variability or dispersion of the categories themselves influences generalization to ambiguous regions of the category space (e.g. the region between the two categories). The general approach is to train participants with examples from a high variability category and a low variability category. Participants are then tested with novel items located within ambiguous regions of the category space which allow the experimenters to assess whether the difference in category variability influenced how far participants generalize the category boundaries. @cohenCategoryVariabilityExemplar2001 conducted two experiments with this basic paradigm. In experiment 1, a low variability category composed of 1 instance was compared against a high-variability category of 2 instances in one condition, and 7 instances in another. In experiment 2 both categories were composed of 3 instances, but for the low-variability group the instances were clustered close to each other, whereas the high-variability groups instances were spread much further apart. Participants were tested on an ambiguous novel instance that was located in between the two trained categories. Both experiments provided evidence that participants were much more likely to categorize the novel middle stimulus into the category with greater variation.

@@

Early models of learning often assumed that discrete experiences with some task or category were not stored individually in memory, but instead promoted the formation of a summary representation, often referred to as a prototype or schema, and that exposure to novel examples would then prompt the retrieval of whichever preexisting prototype was most similar. In addition to being a landmark study on the influence of training variability, @posnerGenesisAbstractIdeas1968 (described above) also put forward an influential argument concerning the nature of the mental representations acquired during learning - namely that learners tend to abstract a prototype, or aggregate representation of the dot pattern categories, rather than encoding each individual stimuli. Recall that participants are trained on only on distortions of the category prototypes (e.g.  low, medium or high distortions), never encountering the exact prototypes during the training stage. Then, in the testing phase, participants are tested with the prototype patterns, their old training items, and novel low, medium and high distortions. The authors found that participants had the highest testing accuracy for the previously unseen prototype patterns, followed by the old training items, and then the novel low, medium and high distortions. The authors interpreted this pattern as evidence that participants had acquired prototype representation of the category, as opposed to storing each individual training instance, and that generalization was based on the similarity of the testing items to the learned prototype representations. @posnerGenesisAbstractIdeas1968 has been extremely influential,  and continues to be cited as evidence that prototype abstraction underlies the benefits of varied training. It's also referenced as a key influence in the development of the "Schema Theory of Motor Learning" @schmidtSchemaTheoryDiscrete1975, which in turn influenced decades of research on the potential benefits of varied training in motor skill learning. However a number of the core assumptions utilized by @posnerGenesisAbstractIdeas1968 were later called into question both empirically and with competing theoretical accounts [@hintzmanMINERVASimulationModel1984; @hintzmanSchemaAbstractionMultipletrace1986; @palmeriCentralTendenciesExtreme2001; @zakiHighdistortionEnhancementEffect2007; @mcclellandDistributedMemoryRepresentation1985; @nosofskyInvestigationsExemplarBasedConnectionist1992; @knappTheoryCategorizationBased1984]. @palmeriCentralTendenciesExtreme2001 demonstrated both the dangers of assuming that psychological representations mimic the metric stimulus space, as well the viability of models with simpler representational assumptions. These authors conducted a near replication of the @posnerGenesisAbstractIdeas1968 study, but also had participants provide similarity judgements of the dot pattern stimuli after completing the training phase. A multidimensional scaling analysis of the similarity judgements revelead that the psychological representations of the prototype stimuli were not located in the middle of the training stimuli, but were instead extreme points in the psychological space. The authors also demonstrated the generalization patterns of @posnerGenesisAbstractIdeas1968 could be accounted for by an exemplar-based model, without any need to assume the abstraction of a prototype.
